Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-09 paper/tidy-missing-data-paper.Rmd
finish improving readability / getting as close to 30 pages as possible.

As data science has become a more solid field, theories and principles have developed to describe best practices. One such idea is 'tidy data,' a set of guidelines for converting raw data into a clean, analysis-ready format that works efficiently in a data analysis pipeline, while another is the idea of the Grammar of Graphics, which describes how to map data values into a visualization [@Wickham2014, Wilkinson2012]. These principles have guided analysts in their work for several years now. However, while these principles outline how to work with data in a general way, they do not address the problem of missing data. In particular, there is little to no guidance for working with missing values in the field of visualization. Most analysis software simply drops missing values when making a plot, although some packages provide a warning (see Figure \@ref(fig:warning)).

@@

Features of tidy data were formally discussed in 2014 [@Wickham2014], subsequently generating much discussion (see @Donoho2017 and commentary, @r4ds), and tools for data analysis. Tidy data is defined as:

@@

\pkg{Amelia} [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior.

@@

\pkg{norm} [@norm], provides multiple imputation using EM for multivariate normal data, drawing from methods in the NORM software [@schafer-norm]. \pkg{norm} does not provide a framework for tracking missing values, instead providing tools for making inference from multiple imputation.

@@

1. Missing value label: A simple label to the shadow matrix clear, "NA" and "!NA" for missing and not missing, represented as 0 and 1 (see Figure \@ref(fig:shadow-matrix-progress) parts 2 and 3 below).

@@

These additions are now discussed in turn.

@@

Other statistical programs represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` through to `.z`. The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state. Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.

@@

A consistent short suffix "`_NA`" for each variable keeps names coordinated throughout analysis (See Figure \@ref(fig:shadow-matrix-progress) part 4). The suffix is short and easy to remember in analysis or visualization. It also makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`. This subtle change is important as it shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.

@@

_Nabular_ data binds the shadow matrix column-wise to the original data, and is so named as a portmanteau of `NA` and `tabular`. _Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, explicitly linking missing values to the data. _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`. _Nabular_ data is a snapshot of the missingness of the data. This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the appropriate, coordinated names. _Nabular_ data is created with `nabular(data)` (see Figure \@ref(fig:nabularfig)). _Nabular_ data is not unlike classical data formats with quality or flag columns associated with each measured variable, e.g. Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

@@

The _nabular_ data structure for missing data provides three key benefits over on-the-fly creating a logical matrix where TRUE means missing, and FALSE means present. Firstly, missing value labels `NA` and `!NA` are clearer. Secondly, special missing values cannot be added fluently during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these value changes.

@@

Common operations on missing data can be considered to be verbs: scan, replace, add, shadow, impute, track, and flag. Missing values can be **scanned** to find possible missings not coded as `NA`. These values can then be **replaced** with an `NA` value. To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data. Data values can be augmented with **shadow** values, which helps explore missing data, as well as facilitating the process of **imputing**, and **tracking**. Finally, unusual or specially coded missing values can be **flagged**.

@@

Summary information such as the number of missings in a given case can be useful in understanding missingness structure, and even more so when this information is kept alongside the data. Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data. `naniar` includes functions to add missingness summary information to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (see Table \@ref(tab:add-missing-info)). There are also functions for adding information about shadow values or nice labels for any missing values with `add_label_shadow()` and `add_label_missings()`.

An example use of these features is in @Tierney2015, where the proportion of
missings is used as the outcome in a model, to identify variables and values
important in predicting missingness structures.

@@

_Nabular_ data has the shadow matrix column-bound to the existing data. This facilitates visualization and summaries, and allows for imputed values to be appropriately tracked. _Nabular_ data can be created with `nabular()`. An example use of `nabular()` some example data set is shown below:

@@

```{r recode-shadow, eval = TRUE, echo = TRUE}
dat_ms %>%
  nabular() %>%
  recode_shadow(x = .where(x == -99 ~ "broken_sensor"))
```

@@

`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`. Instead, _nabular_ data is designed to work with these. `naniar` does provide a few imputation methods to facilitate exploration and visualizations: `impute_mean`, `impute_median`, and `impute_below`. While useful to explore structure in missingness, they are not recommended for use in analysis. `impute_below` imputes values below the minimum value, with some jitter (small random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.

@@

```{r bind-impute-label-example, echo = TRUE}

aq_imputed <- airquality %>%
  nabular() %>%
  simputation::impute_lm(Ozone ~ Temp + Wind) %>%
  simputation::impute_lm(Solar.R ~ Temp + Wind) %>%
  add_label_shadow()

aq_imputed
```

@@

```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values from a linear model. Imputed values are colored green, and data values orange. Imputed values are similar, but slightly trended to the mean of existing data.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}

@@

The number of missings in each variable and case are visualized using `gg_miss_var` and `gg_miss_case` (see Figure \@ref(fig:gg-miss-case-var) below). These are shown using the "airquality" dataset, included in base R, which contains daily air quality measurements in New York, from May to September, 1973. Rather than highlighting the amount of complete data, these visualizations draw attention to the amount of missings, ordering by missingness.

```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in variables (A) and cases (B) for the airquality data. There are missing values in Ozone and Solar.R, with Ozone having more missings. Not many cases have two missing values, with most missingness being from cases with one missing value.", fig.height = 3, fig.width = 6, out.width = "90%", fig.align="center"}

@@

All missings in a dataset can be displayed using a heatmap style visualization. This is achieved using `vis_miss()` from the \pkg{visdat} package [@visdat], which also provides summaries of missingness overall in the legend, and for each column (see Figure \@ref(fig:vis-miss)). The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:vis-miss)B). Similar visualizations are available in other missing data packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation. Using \pkg{ggplot2} means the plot can be customized and combined with other ggplot graphics.

@@

The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017], a visualization technique for showing the size and features of sets in data similar to a venn diagram, but scales better with more variables (Figure \@ref(fig:airquality-upset)). Figure \@ref(fig:airquality-upset) shows only Ozone and Solar.R have missing values. The dots in the bottom right show the combinations of missingness, with the bottom left showing the missingness in each variable, and the top showing the missingness for the combinations. The bottom left shows Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone have missing values together.

```{r airquality-upset, fig.cap = "The pattern of missingness in the airquality dataset shown in an upset plot. Only Ozone and Solar.R have missing values, and Ozone has the most missing values. There are 2 cases where both Solar.R and Ozone have missing values.", fig.height = 3, fig.width = 4, out.width = "80%"}

@@

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed values below the range of data using `impute_below` (by default imputing 10% below the range). Visualizing this as a histogram shows the number of missing values on its left. Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot, shown in Figure \@ref(fig:impute-shift-histogram).

@@

```{r bind-shadow-density, fig.height = 3, fig.width = 5, out.width = "49%", error = FALSE, message = FALSE, fig.cap = "A visualization of temperature according to missingness in ozone from in the airquality dataset. A histogram of temperature facetted by the missingness of ozone (A), or a density of temperature colored by missingness in ozone (B). The distribution shows a cluster of low temperature observations with missing ozone values, but the temperature values are otherwise similar.", fig.show = 'hold'}

@@

To visualize missing values in two dimensions the missing values can be placed in plot margins, by imputing values below the range of the data. As in Section \@ref(univariate-graphics), using _nabular_ data identifies imputed values, and color makes missingness pre-attentive [@treisman1985]. The steps of imputing and coloring are combined into `geom_miss_point()` from the \pkg{naniar} package (Figure \@ref(fig:geom-miss)). There is a mostly uniform spread of missing values for Solar.R and Ozone.

@@

As `geom_miss_point` is a defined geometry for \pkg{ggplot2}, it provides features such as faceting and mapping other variables to aesthetics such as color, shape, or size. The shadow format by itself can also be used to visually explore some patterns of missingness.

@@

Parallel coordinate plots can visualize missingness beyond two dimensions. They transform variables to the same scale, and typically make the data values range between 0 and 1. To showcase this visualization, the dataset `oceanbuoys` from \pkg{naniary} is used, which contains measurements of moored ocean buoys to understand and predict El Niño and El Niña. The data were collected in 1993 and 1997, and contains information on the sea and air temperature, humidity, and east west and north south wind directions.

Figure \@ref(fig:parallel-cord-plot) shows a parallel coordinate plot of the `oceanbuoys` data, with missing values imputed to be 10% below the range, and values colored according to whether humidity contained missing values. Although a 10% below imputation is not the most ideal way to display missing values in a parallel coordinate plot, it shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.

@@

There are different ways to summarize the amount of missing and complete values in a dataset. This section describes approaches to summarizing missingness, implemented in the \pkg{naniar} package. The functions are easy to remember as they follow consistent naming, and their output is consistent, and returns a single number or dataframe. This means they integrate well with plotting and modelling tools. Section \@ref(single-num-sum) discusses single number summaries; Section \@ref(sum-tab-missings) discusses variable- and case-wise summaries; and Section \@ref(num-sum-w-group) shows how the design of these tools works with other tools in an analysis pipeline.

@@

The overall number, proportion, or percent of missing values in a dataset is shown with `n_miss`, `prop_miss` and `pct_miss`. Complementing these are functions to reveal the number of complete values: `n_complete`, `prop_complete`, and `pct_complete`. These can summarize missings for variables and cases by appending `_case` or `_var` to any of the summaries. Table \@ref(tab:n-prop-pct-miss-complete) shows the names and output of these functions on the dataset 'airquality'.

@@

Presenting the number and percent of missing values for each variable, or case, provides a summary usable in models to inform imputations or data handling. For example, potentially dropping variables or deciding to include others in an imputation model. Another useful approach is to tabulate the frequency of missing values for each variable or case; that is, the number of times there are zero, one, two, and so on, missing values. These summaries and tabulations are shown for variable in Tables \@ref(tab:miss-var-summary) and \@ref(tab:miss-var-table), using the `miss_var_summary` and `miss_var_table` functions. Functions are also provided for case-wise summaries and tabulations with `miss_case_summary` and `miss_case_table`, which order by `n_miss`, so the most missings are always shown at the top. Other summaries provided in \pkg{naniar} include summaries of missingness across a provided repeating span, `miss_var_span`, and finding the streaks or runs of missingness in a given variable in `miss_var_run`.

@@

miss_var_summary(airquality) %>% 
  knitr::kable(
    caption = "The \\texttt{miss\\char`_var\\char`_summary} function provides a summary of the number and percent of missings in each variable. Only ozone and solar radiation have missing values.",
    digits = 1)
    # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))

@@

miss_var_table(airquality) %>% 
  knitr::kable(
    caption = "The \\texttt{miss\\char`_var\\char`_table} function provides a tabulation of the amount of missing data in each variable. This shows the number of variables with no missings, 7 missings, and 37 missings, and the percentage of variables with those amounts of missingness. There are not many patterns of missingness.",
    digits = 1)
    # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))
```

@@

These summaries and tabulations can be composed with the "pipe" operator to perform grouped summaries using the `group_by` operator from `dplyr`.
These summaries return dataframes, which may be used by other plotting devices, modelling tools, or other summaries. Table \@ref(tab:group-miss-var-summary) shows an example of the missing data summaries for the airquality dataset in each variable, for each month of the year. These datasets can be summarized in visualizations, or values input into a model.

@@

airquality %>%
  group_by(Month) %>%
  miss_var_summary() %>%
  ungroup() %>%
  slice(1:10) %>%
  knitr::kable(caption = "The \\texttt{miss\\char`_var\\char`_summary} provides a grouped summary of the missingness in each variable, for each Month of the airquality dataset, with only the first 10 rows of output shown. There are more ozone missings in June than May.",
               digits = 1)
               # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

To illustrate the methods, data on housing for the city of Melbourne from January 28, 2016 to March 17, 2018 is used. The data was compiled by scraping weekly property clearance data [@Kaggle-2018-data]. There are `r scales::number(nrow(housing), big.mark = ",")` properties, and `r ncol(housing)` variables in the dataset. The variables include the type of real estate (town house, unit, house), suburb, method of selling, number of rooms, price, real estate agent, date of sale, and the distance from the Central Business District (CBD).

@@

```{r housing-miss-case-var, fig.cap = "The amount of missings in each variable (A) and in each case (B) for Melbourne housing data. (A) Build area and year built have more than 50\\% missing values, and car, bathroom, bedroom2 and longitude and latitude have about 25\\% missings. (B) There are between 5 and 50\\% missing values in cases. There are many missing values in the data, with the majority of missingness being in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}

@@

Possible missingness structures are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)). There are three main clusters of missing data: at the top many variables go missing together; in the middle, building price and year built, at the bottom, building area, year built, and land size.

```{r applic-vis-miss, fig.height = 4, fig.width = 6, out.width = "85%", fig.show='hold', fig.cap = "Heatmap of clustered missingness for the housing data. Three groups of missingness are apparent, at the top for building area to longitude, the middle for building area and year built, at the end for building area, year built, and landsize. There is some structure in the missings.", dev = "png", dpi = 300}

@@

Missingness patterns can also be shown with an `upset` plot [@Conway2017], to display the 8 of intersecting sets of missing variables (see Figure \@ref(fig:applic-vis-miss)). Two patterns stand out: two, and five variables missing, providing further evidence of the missingness patterns seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).

```{r housing-upset, fig.height = 5, fig.width = 8, fig.cap = "An upset plot of eight sets of missingness in the housing data. Two missingness patterns are clear, year built and building area, and lattitude through to building area.", out.width = "100%"}

@@

Tabulating the number of missings in variables (see Table \@ref(tab:housing-miss-var-case-table) (left)) shows three groupings: 6254 - 6824 missings cases in variables, 9265 in another variable, and between 15163 and 16736 missing. Tabulating missings in cases (Table \@ref(tab:housing-miss-var-case-table) (right)) shows 32% of cases have complete data, 26% have 2 missing, 22% have 8 missing, 12% have 1 missing, and 5% have 3 missing, the remaining cases with missings are less than 1%. Two variables with more than 50% missingness are omitted from analysis: building area and year built.

@@

knitr::kable(list(t1, t2),
             caption = "Tabulating missingness for variables (left) and cases (right) to understand missingness patterns. 14 variables have between 0 and 3 missings, then 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missing values. About 30\\% of cases have no missing values, then 45\\% of cases have between 1 and 6 missing values, and then about 23\\% of cases have 8 or more missings. There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.",
             digits = 1)
             # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

Using the information from the overview visualizations and tables, the following variables are explored for features predicting missingness: Land size, latitude, longitude, bedroom2, bathroom, car, and land size.

Missingness structure can be better understood and evaluated by using the data to predict where the data goes missing. This can be achieved by clustering the missing values into groups, and then applying classification and regression trees (CART) to find those variables and values predicting missingness clusters [@Tierney2015; @Barnett2017]. Two clusters are identified, one for each missingness cluster. The missingness clusters are predicted using all variables in the dataset with the CART package `rpart` [@rpart], and plotted using the `rpart.plot` package [@rpart-plot]. 
Importance scores from the CART model describe those variables important for predicting missingness: rooms, price, suburb, council area, distance, and region name. These variables are important for predicting missingness, so are important to include in the imputation model.

@@

```{r rpart-plot, fig.height = 4, fig.width = 8, out.width = "100%", fig.align = "center", fig.cap = "Decision tree output predicting the clusters of missingness. Type of house, the year quarter, and year were important for predicting missingness cluster. The cluster with the most missingness was for quarters 1 and 4, for 2017 and 2018. Type of house, year, and year quarter are important features related to the missingness structure."}

@@

Imputed data are combined by row binding imputation models. The data are reshaped into long format to compare imputed values for each model for 4 variables (Figure \@ref(fig:imputed-by-model)). Compared to KNN imputed values, the linear model imputed values closer to the mean value.

>>>>

As data science has become a more solid field, theories and principles have developed to describe best practices. One such idea is 'tidy data,' a set of guidelines for converting raw data into a clean, analysis-ready format that works efficiently in a data analysis pipeline, while another is the idea of the Grammar of Graphics, which describes how to map data values into a visualization [@Wickham2014, Wilkinson2012]. These principles have guided analysts in their work for several years now. However, while these principles outline how to work with data in a general way, they do not address the problem of missing data. In particular, there is little to no guidance for working with missing values in the field of visualization. Most analysis software simply drops missing values when making a plot, although some packages provide a warning (Figure \@ref(fig:warning)).

@@

Features of tidy data were formally discussed in 2014 [@Wickham2014], subsequently generating much discussion (@Donoho2017 and commentary, @r4ds), and tools for data analysis. Tidy data is defined as:

@@

\pkg{Amelia} [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior. \pkg{norm} [@norm], provides multiple imputation using EM for multivariate normal data, drawing from methods in the NORM software [@schafer-norm]. \pkg{norm} does not provide a framework for tracking missing values, instead providing tools for making inference from multiple imputation.

@@

@@

1. Missing value label: A simple label to the shadow matrix clear, "NA" and "!NA" for missing and not missing, represented as 0 and 1 (Figure \@ref(fig:shadow-matrix-progress) parts 2 and 3).

@@

@@

Other statistical programs represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`. The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state. Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.

@@

A consistent short suffix "`_NA`" for each variable keeps names coordinated throughout analysis (Figure \@ref(fig:shadow-matrix-progress) part 4). The suffix is short and easy to remember in analysis or visualization. It also makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`. This subtle change is important as it shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.

@@

_Nabular_ data binds the shadow matrix column-wise to the original data. It is a portmanteau of `NA` and `tabular`. _Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, explicitly linking missing values to data. _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`. _Nabular_ data is a snapshot of the missingness of the data. This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the coordinated names. _Nabular_ data is created with `nabular(data)` (Figure \@ref(fig:nabularfig)). _Nabular_ data is not unlike classical data formats with quality or flag columns associated with each measured variable, e.g. Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

@@

The _nabular_ data structure provides three key benefits over on-the-fly creating a missingness matrix $R$. Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE. Secondly, special missing values cannot be added easily during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these value changes.

@@

Common missing data operations can be considered verbs: scan, replace, add, shadow, impute, track, and flag. Missing values can be **scanned** to find possible missings not coded as `NA`. These values can then be **replaced** with `NA`. To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data. Data values can be augmented with **shadow** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**. Finally, unusual or specially coded missing values can be **flagged**.

@@

Summary information such as the number of missings in a given case can be useful in understanding missingness structure, and even more so when this information is kept alongside the data. Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data. `naniar` includes functions to add missingness summary information to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)). There are also functions for adding information about shadow values or nice labels for any missing values with `add_label_shadow()` and `add_label_missings()`. An example use of these features is in @Tierney2015, where the proportion of
missings is used as the outcome in a model, to identify variables and values
important in predicting missingness structures.

@@

_Nabular_ data has the shadow matrix column-bound to existing data. This facilitates visualization and summaries, and allows for imputed values to be tracked. _Nabular_ data can be created with `nabular()`, as below:

@@

```{r recode-shadow, eval = TRUE, echo = TRUE}
nabular(dat_ms) %>%
  recode_shadow(x = .where(x == -99 ~ "broken_sensor"))
```

@@

`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation \@ref(imputation). Instead, _nabular_ data is designed to work with existing imeplementation. `naniar` does provide a few imputation methods to facilitate exploration and visualization: `impute_mean`, `impute_median`, and `impute_below`. While useful to explore structure in missingness, they are not recommended for use in analysis. `impute_below` imputes values below the minimum value, with some jitter (small random noise) to reduce overplotting. It has options to change the default amount of shift, and jitter.

@@

```{r bind-impute-label-example, echo = TRUE}
aq_imputed <- nabular(airquality) %>%
  simputation::impute_lm(Ozone ~ Temp + Wind) %>%
  simputation::impute_lm(Solar.R ~ Temp + Wind) %>%
  add_label_shadow()

head(aq_imputed)
```

@@

```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values from a linear model. Imputed values are colored green, and data values orange. Imputed values are similar, but slightly trended to the mean.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}

@@

Rather than highlighting the amount of complete data, missing data visualizations should draw attention to the amount of missings, and ordering by missingness. Overview visualisations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)). These are shown using the "airquality" dataset, included in base R, containing daily air quality measurements in New York, from May to September, 1973.

```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in variables (A) and cases (B) for the airquality data. There are missing values in Ozone and Solar.R, with Ozone having more missings. Not many cases have two missings. Most missingness is from cases with one missing value.", fig.height = 3, fig.width = 6, out.width = "90%", fig.align="center"}

@@

All missings can be displayed using a heatmap style visualization. This is achieved using `vis_miss()` from the \pkg{visdat} package [@visdat], which also provides summaries of missingness overall in the legend, and for each column (Figure \@ref(fig:vis-miss)). The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:vis-miss)B). Similar visualizations are available in other missing data packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation. Using \pkg{ggplot2} makes the plot easily customizable.

@@

The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017]. Similar to a Venn diagram, an upset plot shows the size and features of overlapping sets, but scales well with more variables. Figure \@ref(fig:airquality-upset) shows only Ozone and Solar.R have missing values. The bottom right show the combinations of missingness, the bottom left showing missingness in each variable, and the top showing the combinations of missingness. The bottom left shows Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone go missing together.

```{r airquality-upset, fig.cap = "The pattern of missingness in the airquality dataset shown in an upset plot. Only Ozone and Solar.R have missing values, and Ozone has the most missing values. There are 2 cases where both Solar.R and Ozone have missing values.", fig.height = 3, fig.width = 4, out.width = "75%"}

@@

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed values below the range of data using `impute_below` (by default imputing 10% below the range). Figure \@ref(fig:impute-shift-histogram) shows a histogram of Ozone, the number of missing values on its left. Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot.

@@

```{r bind-shadow-density, fig.height = 3, fig.width = 5, out.width = "49%", error = FALSE, message = FALSE, fig.cap = "Temperature according to missingness in ozone from in the airquality dataset. A histogram of temperature facetted by the missingness of ozone (A), or a density of temperature colored by missingness in ozone (B). These show a cluster of low temperature observations with missing ozone values, but temperature is otherwise similar.", fig.show = 'hold'}

@@

To visualize missing values in two dimensions the missing values can be placed in plot margins, by imputing values below the range of the data.  Using _nabular_ data identifies imputed values (\@ref(univariate-graphics)), and color makes missingness pre-attentive [@treisman1985]. The steps of imputing and coloring are combined into `geom_miss_point()`. Figure \@ref(fig:geom-miss) shows a mostly uniform spread of missing values for Solar.R and Ozone. As `geom_miss_point()` is a defined \pkg{ggplot2} geometry, it works with features such as facetting and mapping other variables to graphical aesthetics. The shadow format by itself is also useful for exploring patterns of missingness.

@@

@@

Parallel coordinate plots can visualize missingness beyond two dimensions. They transform variables to the same scale, ranging between 0 and 1. The `oceanbuoys` dataset from \pkg{naniar} is used for this visualisation, containing measurements of ocean buoys to understand and predict El Niño and El Niña. Data was collected in 1993 and 1997, and contains measurements of sea and air temperature, humidity, and east west and north south wind speeds. Figure \@ref(fig:parallel-cord-plot) shows a parallel coordinate plot of the `oceanbuoys`, with missing values imputed to be 10% below the range, and values colored according to where humidity was missing (humidity_NA). This shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.

@@

This section describes approaches to summarizing missingness. The functions are implemented in \pkg{naniar}, and are easy to remember as they have consistent names and output, returning either a single number \@ref(single-num-sum), or a dataframe (\@ref(sum-tab-missings)), integrating well with plotting and modelling tools. How these work with other tools in an analysis pipeline is shown in \@ref(num-sum-w-group).

@@

The overall number, proportion, or percent of missing values in a dataset is shown with `n_miss`, `prop_miss` and `pct_miss`, and their complete value complements: `n_complete`, `prop_complete`, and `pct_complete`. Summaries for variables and cases are made by appending `_case` or `_var` to any of the summaries. An overview is shown in Table \@ref(tab:n-prop-pct-miss-complete).

@@

Presenting the number and percent of missing values for each variable, or case, provides a summary usable in models to inform imputations or data handling. For example, potentially dropping variables or deciding to include others in an imputation model. Another useful approach is to tabulate the frequency of missing values for each variable or case; that is, the number of times there are zero, one, two, and so on, missing values. These summaries and tabulations are shown for variable in Tables \@ref(tab:miss-var-summary) and \@ref(tab:miss-var-table), using `miss_var_summary` and `miss_var_table`. Functions are also provided for case-wise summaries and tabulations: `miss_case_summary` and `miss_case_table`, which order rows by `n_miss`, to show the most missings at the top. `naniar` also shows missingness summaries across a repeating span, `miss_var_span`, and finding streaks or runs of missingness in a given variable in `miss_var_run`.

@@

miss_var_summary(airquality) %>% 
  knitr::kable(
    caption = "\\texttt{miss\\char`_var\\char`_summary} provides the number and percent of missings in each variable in airquality. Only ozone and solar radiation have missing values.",
    digits = 1)
    # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))

@@

miss_var_table(airquality) %>% 
  knitr::kable(
    caption = "\\texttt{miss\\char`_var\\char`_table} tabulates the amount of missing data in each variable in airquality. This shows the number of variables with 0, 7, and 37 missings, and the percentage of variables with those amounts of missingness. There are few missingness patterns.",
    digits = 1)
    # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))
```

@@

These summaries and tabulations can be composed with the "pipe" operator to perform grouped summaries with `dplyr`'s `group_by` function.
These summaries return dataframes, easily used in other plotting devices, modelling tools, or other summaries. Table \@ref(tab:group-miss-var-summary) shows an example of missing data summaries for airquality by each month.

@@

airquality %>%
  group_by(Month) %>%
  miss_var_summary() %>%
  ungroup() %>%
  slice(1:10) %>%
  knitr::kable(caption = "\\texttt{miss\\char`_var\\char`_summary} combined with \\texttt{group\\char`_by} provides a grouped summary of the missingness in each variable, for each Month of the airquality dataset. Only the first 10 rows are shown. There are more ozone missings in June than May.",
               digits = 1)
               # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

To illustrate the methods, data on housing for the city of Melbourne from January 28, 2016 to March 17, 2018 is used. The data was compiled by scraping weekly property clearance data [@Kaggle-2018-data]. There are `r scales::number(nrow(housing), big.mark = ",")` properties, and `r ncol(housing)` variables in the dataset. The variables include real estate type (town house, unit, house), suburb, selling method, number of rooms, price, real estate agent, sale date, and distance from Central Business District (CBD).

@@

```{r housing-miss-case-var, fig.cap = "The amount of missings in variables (A) and cases (B) for Melbourne housing data. (A) Build area and year built have more than 50\\% missing, and car, bathroom, bedroom2 and longitude and latitude have about 25\\% missings. (B) Cases are missing 5 - 50\\% of values. The majority of missingness is in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}

@@

Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).

```{r applic-vis-miss, fig.height = 4, fig.width = 6, out.width = "85%", fig.show='hold', fig.cap = "Heatmap of clustered missingness for housing data reveals structured missingness. Three groups of missingness are apparent. At the top: building area to longitude; the middle: building area and year built; the end: building area, year built, and landsize.", dev = "png", dpi = 300}

@@

Figure \@ref(fig:applic-vis-miss) shows missingness patterns with an `upset` plot [@Conway2017], displaying 8 intersecting sets of missing variables. Two patterns stand out: two, and five variables missing, providing further evidence of the missingness patterns seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).

```{r housing-upset, fig.height = 5, fig.width = 8, fig.cap = "An upset plot of 8 sets of missingness in the housing data. Missingness for each variable is shown on the bottom left. Connected dots show co-occurences of missings in variables. Two missingness patterns are clear, year built and building area, and lattitude through to building area.", out.width = "100%"}

@@

Tabulating the number of missings in variables in Table \@ref(tab:housing-miss-var-case-table) (left) shows three groups of missingness. Tabulating missings in cases (Table \@ref(tab:housing-miss-var-case-table) (right) shows six patterns of missingness. These overview plots lead to the removal of two variables from with more than 50% missingness from analysis: building area and year built.

@@

knitr::kable(list(t1, t2),
             caption = "Tabulating missingness for variables (left) and cases (right) to understand missingness patterns. 14 variables have 0 - 3 missings, 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missings. About 30\\% of cases have no missings, 45\\% of cases have 1 - 6 missings, and about 23\\% of cases have 8 or more missings. There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.",
             digits = 1)
             # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

Using information from \@ref(case-study-explore-pattern), the following variables are explored for features predicting missingness: Land size, latitude, longitude, bedroom2, bathroom, car, and land size.

Missingness structure is explored by clustering the missing values into groups. Then, a  classification and regression trees (CART) model is applied to predict these missingness clusters using the remaining variables, to find variables and values that predicting missingness clusters [@Tierney2015; @Barnett2017]. Two clusters of missingness are identified and predicted using all variables in the dataset with the CART package `rpart` [@rpart], and plotted using the `rpart.plot` package [@rpart-plot].  Importance scores reveal variables important for predicting missingness: rooms, price, suburb, council area, distance, and region name. These variables are important for predicting missingness, so are important to include in the imputation model.

@@

```{r rpart-plot, fig.height = 4, fig.width = 8, out.width = "90%", fig.align = "center", fig.cap = "Decision tree output predicting missingness clusters. Type of house, year quarter, and year were important for predicting missingness cluster. The cluster with the most missingness was for quarters 1 and 4, for 2017 and 2018. Type of house, year, and year quarter are important features related to missingness structure."}

@@
